\section{Discussion}
\label{sec:discussion}

\subsection{What the results actually say}

The two cleanest takeaways from the empirical section are these. First, the rolling intermarket dependency between Bitcoin and conventional assets is genuinely forecastable out of sample---not a fluke of in-sample overfitting, but a real signal that survives strict temporal ordering constraints across all six pairs and four window lengths. Second, that forecastability is almost entirely driven by persistence. Rolling correlation does not jump around randomly; it remembers yesterday.

Both findings matter, and neither cancels the other out.

\subsection{Why AR1 wins, and what that actually means}

The dominance of simple autoregressive baselines is not embarrassing. It tells you something specific about the data-generating process. A rolling Pearson correlation computed over $w$ days shares $w-1$ observations with the previous window, which mechanically creates strong autocorrelation. After Fisher transformation, the first-order lag captures the overwhelming majority of forecastable variance. Under those conditions, any sensible model will lean heavily on the most recent value---and a model that does nothing else will be very hard to beat.

This is a documented pattern in financial forecasting \cite{welch2008}. Signal-to-noise ratios in financial markets are low. The incremental information in auxiliary features is often swamped by the autoregressive component. Recognising this early would have saved quite a bit of time running hyperparameter search on Random Forest configurations.

That said, AR1 winning does not mean machine learning is useless here. It means machine learning is useful for a different reason: ruling out more flexible alternatives. The fact that a well-tuned XGBoost model cannot beat AR1 is itself informative. It tells you the dependency process does not contain exploitable nonlinear structure at the one-step horizon, at least not in the feature set used here.

\subsection{ML versus DCC-GARCH: a fairer comparison}

Ridge and ElasticNet consistently beat DCC-GARCH. That gap is large---average RMSE of around 0.09 versus 0.21. The DM statistics are all positive with $p$-values at the numerical floor of double precision.

The right interpretation is not that DCC-GARCH is a bad model. It is a theoretically coherent model for conditional covariance that was designed for a different job. Its output is an instantaneous conditional correlation, not a smoothed rolling average. Projecting one onto the other introduces a systematic lag that the model was never designed to eliminate. Machine-learning models, trained directly on the Fisher-transformed rolling target, do not have that mismatch. They are optimised for the right objective from the start.

This is the honest explanation for the performance gap, and it should be stated clearly in the thesis rather than celebrated as if ML had somehow defeated an econometric competitor on its home turf.

\subsection{The investor signal: what works and what does not}

The signal layer produces results that are moderate in every dimension: AUC ranging from roughly 0.46 to 0.55, balanced accuracy ranging from below 0.50 to around 0.545 depending on pair and window, exit rates between 15 and 35\% depending on the model and threshold. None of those numbers is impressive in isolation. Combined, they describe a signal that consistently picks the right direction more often than chance, fires at a reasonable frequency, and is more informative for equity stress than for precious-metals or dollar-index stress.

The equity result is economically interpretable. Bitcoin and the S\&P~500 share exposure to global risk appetite, funding liquidity, and sentiment cycles. When the dependency between them is rising or shifting, it often reflects broader market-regime transitions that also elevate equity downside risk. The crypto feature set is reasonably well-specified for that channel.

Gold and silver respond to a wider set of forces---real rates, inflation expectations, geopolitical premia, safe-haven demand---that are not well represented by Bitcoin dynamics. The attenuated signal for those instruments is therefore not a surprise; it is consistent with a story where the feature set captures some drivers of market stress but not all of them.

In practice, the signal should be treated as an auxiliary risk indicator, not an alarm system. It is best used alongside fundamental valuation, macro analysis, and position-sizing rules---raising the threshold for accepting risk when the dependency-based probability of a stress regime is elevated.

\subsection{Methodological choices that mattered}

A few design decisions turned out to be important:

\begin{itemize}
    \item \textbf{Walk-forward evaluation for DCC-GARCH as well.} Running DCC-GARCH on the full sample before producing out-of-sample predictions would have given it an unfair advantage. Implementing the same expanding-window protocol for all models leveled the comparison.
    \item \textbf{Fisher transformation on the target.} Without it, the bounded nature of Pearson correlation near $\pm 1$ distorts errors near the extremes. Fisher-$z$ brings the target closer to Gaussian and makes RMSE a more sensible loss function across the full range.
    \item \textbf{Multiple windows.} Using only the 30-day window would have understated how strongly predictability depends on the autocorrelation structure. The progression from $w=14$ to $w=90$ shows the autocorrelation effect systematically.
    \item \textbf{Multiple metrics.} Relying on RMSE alone would have missed that DCC-GARCH is not just worse on average---it is worse for every pair, at every window, with statistical significance. The DM test adds that layer of certainty.
\end{itemize}

\subsection{What the results do not say}

A few things this thesis cannot claim, and should not:

Bitcoin does not predict equity returns. It predicts (partially) the evolving \emph{dependency} between Bitcoin and equities, which is a weaker and more precise statement. The signal layer predicts elevated downside risk probability for conventional assets---not direction, not magnitude, and not timing within a flagged period.

The results are based on one-step-ahead forecasting. Multi-step performance could differ substantially, especially for machine learning where the autoregressive advantage weakens as the horizon extends. This is an open question.

Finally, no transaction-cost analysis was done. Whether the signal produces risk-adjusted gains in a real portfolio after costs and turnover constraints is unknown. That would require a separate study with explicit portfolio construction assumptions.

\subsection{Iterative model development: what changed across versions}
\label{subsec:iterations}

The final model presented in this thesis is the result of several months of iterative refinement.
This section documents that process honestly, because the sequence of changes --- what was tried,
what failed, and what eventually helped --- is itself an empirical result about the nature of the
forecasting problem.

\paragraph{Version 1 (March 2026): minimal baseline.}
The first working prototype contained seven models: Naive\_Last, AR1, Ridge, ElasticNet,
Random Forest, Gradient Boosting, and XGBoost.
Feature engineering was minimal: four autoregressive lags of the rolling correlation,
short-window volatilities for both assets, and lagged returns up to lag~5.
No DCC-GARCH benchmark was included at this stage.
Hyperparameter tuning was ad hoc: Ridge $\alpha = 1.0$, ElasticNet $\alpha = 0.01$,
XGBoost with 1000 estimators and learning rate 0.05.
No feature scaling was applied to linear models.
The best OOS RMSE for BTC-USD vs S\&P~500 at $w = 30$ was Ridge at 0.0897,
with AR1 performing comparably.

\paragraph{Version 2 (October 2026): DCC-GARCH and signal layer.}
The second major milestone introduced the DCC-GARCH econometric benchmark via a
walk-forward protocol that matched the expanding-window setup used for machine-learning models.
This was a critical correctness fix: an earlier full-sample DCC fit would have granted the
benchmark access to future data, invalidating the comparison.
The investor signal layer was added as a second stage, classifying whether conventional
assets were entering stress regimes using the rolling-dependency features.
Core model hyperparameters and feature sets were unchanged from Version~1.

\paragraph{Version 3 (November 2026, pre-Turkey checkpoint): stabilisation.}
This version stabilised the pipeline before a planned travel break.
No metric improvements were achieved; the focus was on robustness:
XGBoost GPU-to-CPU fallback, better exception handling, and consistent metric exports.
OOS results were nearly identical to Version~2, confirming that the Version~1 performance
plateau was not an artefact of implementation bugs.

\paragraph{Version 4 (May 2026): feature expansion, regularisation revision, and HAR.}
The final version incorporated several targeted improvements informed by the
Version~1--3 plateau:

\begin{itemize}
    \item \textbf{Feature engineering expanded from $\approx$15 to $\approx$35 features.}
          New features include momentum and z-score of the rolling correlation
          (\texttt{dep\_mom\_5}, \texttt{dep\_mom\_20}, \texttt{dep\_zscore}),
          extended lags (\texttt{dep\_lag20}, \texttt{dep\_lag60}), a 60-day volatility
          window, the volatility ratio between assets, squared-return proxies, and a
          short-run correlation momentum term.
    \item \textbf{HAR model added.}
          The Heterogeneous AutoRegressive (HAR) model
          \cite{corsi2009har}
          was introduced as a theoretically grounded alternative to AR1.
          HAR uses three temporal scales simultaneously:
          $\hat{\rho}_{t+1} = \alpha + \beta_d\,\rho_t + \beta_w\,\bar{\rho}^{(5)}_t + \beta_m\,\bar{\rho}^{(22)}_t + \varepsilon_{t+1}$,
          where $\bar{\rho}^{(k)}_t$ denotes the equally-weighted average of the previous
          $k$ observations.
          Originally developed for realised variance, HAR is well-suited to rolling
          correlations because the overlapping-window structure creates natural
          heterogeneous memory at daily, weekly, and monthly horizons.
    \item \textbf{Regularisation philosophy revised.}
          Ridge regularisation was reduced from $\alpha = 1.0$ to $\alpha = 0.5$
          and an explicit \texttt{StandardScaler} was added.
          The combined effect was dramatic: Ridge RMSE at BTC-USD vs S\&P~500 $w = 30$
          improved from 0.0897 (Version~1) to 0.0651, nearly matching AR1.
          Conversely, XGBoost regularisation was \emph{tightened} (learning rate
          0.05~$\to$~0.02, added $L_1/L_2$ penalties), reflecting the insight that
          the expanded feature set increased overfitting risk for tree-based methods.
    \item \textbf{Adaptive ensemble.}
          The walk-forward ensemble now weights each constituent model by the inverse of
          its RMSE over the most recent 60 out-of-sample predictions, rather than using
          equal weights.
          This allows the ensemble to adapt dynamically to regime shifts in which
          different models temporarily dominate.
    \item \textbf{Bootstrap confidence intervals and sensitivity analyses.}
          500-replicate bootstrap CIs were added for RMSE and $R^2$,
          and a refit-frequency sensitivity sweep across
          $\{5, 10, 21, 42, 63\}$ days was included to verify that the
          $\texttt{refit\_every} = 20$ default is not a cherry-picked choice.
\end{itemize}

\paragraph{What the iterations actually taught.}
The most important lesson from four versions of the same model is negative:
\emph{expanding the feature set from 15 to 35 variables did not help tree-based models}.
XGBoost RMSE was virtually unchanged or slightly worse after the expansion, while Ridge,
which uses the full feature set linearly with regularisation, improved substantially.
This is consistent with the central finding of the thesis: the forecasting target has
a nearly linear autoregressive structure, and the incremental information in
volatility ratios, momentum, or return squares is captured well by a regularised linear
model but does not add exploitable nonlinearity for gradient-boosted trees.

The practical implication for future work is that \emph{better features} are unlikely to
close the remaining gap between AR1 and machine learning on this task.
Improvements would more plausibly come from a fundamentally different modelling approach
--- multi-step horizons, high-frequency data, or joint modelling of multiple correlation pairs
simultaneously.

% ─────────────────────────────────────────────────────────────────────────────
\subsection{Limitations, honestly stated}

The main limitations are: daily frequency (no intraday), one-step horizon, Pearson correlation as the only dependency measure, a single base cryptocurrency (Bitcoin), and no portfolio backtest. Each represents a natural extension that could sharpen or complicate the findings. None of them invalidates the core empirical results within the scope of what was actually tested.
