\section{Conclusion}

This thesis set out to answer a specific question: does information about the evolving relationship between Bitcoin and conventional assets help forecast that relationship, and can those forecasts be converted into something useful for risk management? The short answer is yes to both---with important qualifications on what ``useful'' means.

The dependency process is forecastable out of sample. That result holds across six asset pairs, four rolling window lengths, and multiple model families evaluated under a strict walk-forward protocol. The longer the window, the smoother and more autocorrelated the target becomes, and the higher the out-of-sample $R^2$. At the 90-day window, even the simplest AR(1) model reaches $R^2 > 0.98$ for several pairs.

That last observation is also the central limitation. The forecasting problem is largely solved by persistence. Because a 30-day rolling correlation shares 29 of its 30 data points with the previous day's value, it is hard to beat a model that simply says ``tomorrow's correlation will look a lot like today's.'' Ridge, AR(1), and HAR form a statistically indistinguishable top cluster; the naive last-value baseline sits just behind. Random Forest, GBM, and XGBoost perform substantially worse. Machine learning in its tree-based forms does not overcome the autoregressive baseline---but Ridge does match it, and all ML models reliably outperform DCC-GARCH, which is the theoretically motivated econometric benchmark. That comparison is informative: flexible data-driven methods are better than the parametric alternative at this specific task, even when they cannot beat simple persistence.

The investor signal layer is more nuanced. Balanced accuracy of 50--54\% and AUC values of 0.50--0.55 are modest; results are above chance for most but not all configurations, with the Bitcoin--NASDAQ pair performing below chance at longer windows. Signal quality is strongest for the Bitcoin--S\&P~500 pair, where the crypto feature set aligns best with the macro-financial forces that drive broad equity stress. For gold, silver, and the dollar index, the signal is weaker---which makes economic sense, since those markets respond to drivers that Bitcoin dynamics do not capture well.

The honest practical conclusion: this kind of signal should be treated as one additional input to a risk-management process, not as a standalone alarm. It does not predict tomorrow's returns. It does give a statistically consistent nudge toward caution when the dependency structure is shifting---and that is a defensible use case.

Future work should test the framework at longer forecast horizons (where persistence advantages may decay faster), with richer feature sets including on-chain metrics and macro-financial variables, and against tail-dependence measures that capture joint extreme behaviour more precisely than rolling Pearson correlation. A portfolio backtest with explicit transaction costs would close the gap between statistical performance and economic performance. Those extensions are left open; this thesis provides the foundation.
