\section{Appendix E: Additional Statistical Tests}

This appendix presents three supplementary hypothesis tests that further
validate the robustness of the empirical results: a Mincer-Zarnowitz forecast
efficiency test, residual diagnostic tests (Ljung-Box and ARCH LM), and a
Harvey-Leybourne-Newbold small-sample correction of the Diebold-Mariano statistic.

% ─────────────────────────────────────────────────────────────────────────────
\subsection{Mincer-Zarnowitz Forecast Efficiency Test}

The Mincer-Zarnowitz (MZ) test \cite{mincer1969economic} checks whether forecasts
are unbiased and informationally efficient.
For each model the auxiliary regression
\begin{equation}
    y_t = \alpha + \beta\,\hat{y}_t + \varepsilon_t
\end{equation}
is estimated by OLS and the joint null hypothesis $H_0{:}\;\alpha = 0,\;\beta = 1$
is tested with an $F$-statistic.
Rejection ($p < 0.05$) indicates that the forecast is either biased ($\alpha \neq 0$)
or fails to make full use of available information ($\beta \neq 1$).

Table~\ref{tab:mz_summary} summarises the percentage of the 24
dependency–window experiments in which each model passes the MZ test,
together with its average $\alpha$ and $\beta$ coefficients.

\begin{table}[H]
    \centering
    \caption{Mincer-Zarnowitz forecast efficiency: percentage of 24 experiments
             passing the joint $F$-test ($p \geq 0.05$) and average coefficient estimates.}
    \label{tab:mz_summary}
    \begin{tabular}{lrrrr}
        \toprule
        Model & Efficient (\%) & $\bar{\alpha}$ & $\bar{\beta}$ & $\bar{F}$ \\
        \midrule
        AR1         & 70.8 & $+0.0006$ & 0.9976 &   2.0 \\
        HAR         & 70.8 & $+0.0006$ & 0.9974 &   2.4 \\
        Ridge       & 45.8 & $-0.0015$ & 0.9992 &   4.1 \\
        RF          & 16.7 & $-0.0019$ & 1.0140 &  26.2 \\
        Naive\_Last &  8.3 & $+0.0097$ & 0.9706 &  16.3 \\
        GBM         &  8.3 & $-0.0046$ & 1.0226 &  35.0 \\
        ElasticNet  &  4.2 & $-0.0056$ & 1.0175 &  20.9 \\
        Ensemble    &  0.0 & $-0.0077$ & 1.0252 &  30.1 \\
        XGB\_GPU    &  0.0 & $-0.0252$ & 1.0979 & 202.2 \\
        DCC\_GARCH  &  0.0 & $-0.0881$ & 1.4433 & 237.3 \\
        \bottomrule
    \end{tabular}
\end{table}

AR1 and HAR pass the efficiency test at the same rate (70.8\% of experiments),
confirming that both models make near-optimal use of the available information.
Ridge is the most efficient among the regularised ML models at 45.8\%.
DCC-GARCH never passes: its average $\bar{\beta} = 1.44$ reflects a strong
systematic tendency to over-predict correlation magnitudes, consistent with
the large positive RMSE penalty observed in the main results.

Table~\ref{tab:mz_gspc_w30} provides the full detail for the representative
pair BTC-USD vs S\&P~500 at $w = 30$.

\begin{table}[H]
    \centering
    \caption{Mincer-Zarnowitz test: BTC-USD vs S\&P~500, $w = 30$.
             $F$-statistic tests $H_0{:}\;\alpha=0, \beta=1$; $p_\text{MZ}$ is
             the associated $p$-value.}
    \label{tab:mz_gspc_w30}
    \begin{tabular}{lrrrrl}
        \toprule
        Model & $\hat{\alpha}$ & $\hat{\beta}$ & $F$ & $p_\text{MZ}$ & Efficient \\
        \midrule
        Ridge       & $+0.0071$ & 0.9903 &   4.76 & 0.009 & No  \\
        AR1         & $+0.0071$ & 0.9909 &   5.07 & 0.006 & No  \\
        HAR         & $+0.0074$ & 0.9913 &   5.77 & 0.003 & No  \\
        Naive\_Last & $+0.0114$ & 0.9695 &  16.74 & $<0.001$ & No  \\
        ElasticNet  & $+0.0065$ & 1.0048 &  17.18 & $<0.001$ & No  \\
        Ensemble    & $+0.0086$ & 1.0113 &  39.91 & $<0.001$ & No  \\
        RF          & $+0.0221$ & 0.9951 &  57.04 & $<0.001$ & No  \\
        GBM         & $+0.0226$ & 0.9945 &  62.30 & $<0.001$ & No  \\
        DCC\_GARCH  & $-0.0886$ & 1.4616 & 261.03 & $<0.001$ & No  \\
        XGB\_GPU    & $+0.0436$ & 1.0316 & 318.18 & $<0.001$ & No  \\
        \bottomrule
    \end{tabular}
\end{table}

At this particular pair and window all models are technically rejected, but the
performance gradient is steep. Ridge ($F = 4.76$), AR1 ($F = 5.07$), and HAR
($F = 5.77$) are rejected at the boundary of significance and have coefficients
very close to $(0, 1)$, indicating near-efficient use of information. The
DCC-GARCH statistic (261.0) and XGBoost statistic (318.2) stand apart as the
most severely misspecified, confirming the main-text findings.

% ─────────────────────────────────────────────────────────────────────────────
\subsection{Residual Diagnostics: Ljung-Box and ARCH LM Tests}

Two classical residual tests are applied to the OOS forecast errors
$e_t = y_t - \hat{y}_t$ for every model $\times$ dependency $\times$ window
combination (192 cases in total).

\textbf{Ljung-Box Q-test} (10 lags): tests $H_0{:}$ no serial autocorrelation.
Rejection indicates that the model has not fully captured the temporal structure
of the target series.

\textbf{ARCH LM test} (5 lags): regresses $e_t^2$ on its own lags and tests
$H_0{:}$ no conditional heteroscedasticity.
Rejection signals remaining volatility clustering in the residuals.

Table~\ref{tab:resid_summary} reports, for each model, the percentage of
experiments in which the residuals pass each test at the 5\% level.

\begin{table}[H]
    \centering
    \caption{Residual diagnostics: percentage of 24 experiments passing the
             Ljung-Box autocorrelation test and the ARCH LM heteroscedasticity
             test at $\alpha = 0.05$.}
    \label{tab:resid_summary}
    \begin{tabular}{lrr}
        \toprule
        Model & No autocorr (\%) & No ARCH (\%) \\
        \midrule
        Naive\_Last & 50.0 & 58.3 \\
        AR1         & 41.7 & 62.5 \\
        HAR         & 41.7 & 58.3 \\
        Ridge       & 20.8 & 54.2 \\
        ElasticNet  &  0.0 & 37.5 \\
        Ensemble    &  0.0 & 12.5 \\
        RF          &  0.0 &  0.0 \\
        GBM         &  0.0 &  0.0 \\
        XGB\_GPU    &  0.0 &  0.0 \\
        DCC\_GARCH  &  0.0 &  0.0 \\
        \bottomrule
    \end{tabular}
\end{table}

The autocorrelation results follow a clear gradient. Naive\_Last and the
autoregressive models (AR1, HAR) partially capture the serial structure,
passing the Ljung-Box test in 42--50\% of experiments. Ridge also passes in
21\% of cases, reflecting its ability to exploit autoregressive signal from
the lag features. ElasticNet, despite its linear form, passes in 0\% of
autocorrelation tests but retains meaningful ARCH pass rates (38\%), suggesting
it removes some but not all temporal dependence. Tree-based methods (RF, GBM,
XGBoost) and DCC-GARCH leave strongly significant autocorrelation and
heteroscedasticity in every experiment, confirming that they do not encode
the autoregressive structure of the target and instead produce residuals
that are well-structured and exploitable. The large Q-statistics for
tree-ensemble methods (e.g.\ $Q > 1{,}000$ for XGB\_GPU across most pairs)
underline that tree-based forecasters fail to generalise the AR structure
learned at training time.

% ─────────────────────────────────────────────────────────────────────────────
\subsection{Harvey-Leybourne-Newbold Corrected Diebold-Mariano Test}

The standard Diebold-Mariano statistic uses the asymptotic standard-normal
distribution, which tends to over-reject in finite samples.
Harvey, Leybourne \& Newbold \cite{harvey1997testing} propose multiplying
the DM statistic by the correction factor
\begin{equation}
    k = \sqrt{\frac{n + 1 - 2h + h(h-1)/n}{n}},
\end{equation}
where $n$ is the number of forecast observations and $h$ the forecast horizon,
and comparing the result to a $t(n-1)$ distribution.
For $h = 1$ and $n > 1{,}000$ the factor $k \approx 1.000$, so the practical
effect on $p$-values is negligible.

Applied to all 336 DM comparisons (168 ML models vs DCC-GARCH and 168 vs
Naive\_Last, Table~\ref{tab:dm_hln}) the HLN correction produces zero sign
changes in significance at the 5\% level (average correction factor
$k = 1.000226$).
All conclusions regarding the relative ranking of models, and in particular
the finding that every ML model significantly outperforms DCC-GARCH, are
therefore robust to the small-sample correction.

\begin{table}[H]
    \centering
    \caption{HLN-corrected DM test: ML models vs DCC-GARCH benchmark,
             BTC-USD vs S\&P~500, $w = 30$.
             DM: original statistic (normal); DM$_\text{HLN}$: corrected
             statistic ($t_{n-1}$).
             All models beat DCC-GARCH at $p < 0.001$ under both tests.}
    \label{tab:dm_hln}
    \begin{tabular}{lrrrrl}
        \toprule
        Model & DM & $p_\text{DM}$ & DM$_\text{HLN}$ & $p_\text{HLN}$ & Sig. \\
        \midrule
        Ridge      & 26.86 & $<0.001$ & 26.86 & $<0.001$ & *** \\
        HAR        & 26.83 & $<0.001$ & 26.83 & $<0.001$ & *** \\
        ElasticNet & 26.97 & $<0.001$ & 26.97 & $<0.001$ & *** \\
        Ensemble   & 27.05 & $<0.001$ & 27.04 & $<0.001$ & *** \\
        RF         & 26.44 & $<0.001$ & 26.43 & $<0.001$ & *** \\
        GBM        & 26.16 & $<0.001$ & 26.16 & $<0.001$ & *** \\
        XGB\_GPU   & 24.29 & $<0.001$ & 24.28 & $<0.001$ & *** \\
        \bottomrule
    \end{tabular}
\end{table}

% ─────────────────────────────────────────────────────────────────────────────
\subsection{Automated Validation Suite}
\label{subsec:auto_tests}

In addition to the statistical tests above, the full project ships with an
automated \texttt{pytest} suite of \textbf{81 unit and integration tests}
organised into ten test classes.
The suite is executed automatically as part of the reproducibility runner
(\texttt{python run\_all.py}) between the notebook step and the final
\LaTeX\ compilation, so every rebuilt PDF contains a fresh test report.

The test classes cover the following aspects of the pipeline:
\begin{itemize}
    \item \textbf{Dataset Integrity} --- raw price and return files exist,
          have correct tickers, positive prices, sorted and non-duplicate dates,
          and extend to at least 2026.
    \item \textbf{Data Quality} --- missing-rate thresholds, zero-price
          detection, reasonable annualised volatility, pairwise coverage overlap.
    \item \textbf{Feature Engineering} --- Fisher-$z$ round-trip identity,
          rolling-correlation range, no look-ahead in prediction CSVs.
    \item \textbf{Model Metrics} --- RMSE $>0$, $R^2 \leq 1$, MAE $\leq$ RMSE,
          all windows and expected models present, AR1 beats DCC-GARCH on average.
    \item \textbf{DM Tests} --- finite DM statistics, $p$-values in $[0,1]$,
          correct sign convention for Ridge vs DCC-GARCH.
    \item \textbf{Signal Metrics} --- balanced accuracy and F1 in $[0,1]$,
          AUC above random, exit rate within a sensible range.
    \item \textbf{Walk-Forward} --- correct number of prediction CSVs,
          non-trivial predictions, index alignment with returns.
    \item \textbf{Output Files} --- all expected CSV, JSON, PNG, and \texttt{.tex}
          artefacts are present after a full pipeline run.
    \item \textbf{Bootstrap CI} --- lower $\leq$ mean $\leq$ upper for RMSE and
          $R^2$; CI width below 0.10.
    \item \textbf{Sensitivity} --- all five \texttt{refit\_every} values present;
          RMSE in a physically plausible range.
\end{itemize}

Table~\ref{tab:test_report} lists each individual test together with its
result and wall-clock execution time.
All 81 tests pass on the present outputs; skipped tests indicate that the
optional \texttt{arch} package (DCC-GARCH) is not installed in the current
environment.

\needspace{4\baselineskip}

\IfFileExists{../tables/test_report_auto.tex}{%
    \input{../tables/test_report_auto.tex}%
}{%
    \begin{table}[H]
        \centering
        \caption{Automated validation suite results}
        \label{tab:test_report}
        \small Run \texttt{python run\_all.py} to regenerate this table.
    \end{table}%
}

\clearpage
